Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-01-05 manuscript/manuscript.Rmd
revision 1

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses variability that is more likely to be intragenomic, while maintaining variability that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  A total of 87 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes of the north and south shores of St. John were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave regime. These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [xxx] and wave regime [xxx]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition (xxx). SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite [xxx]. From the 12 monthly climatological values obtained for each sampling year (2003-2010, n=96 months), the median, mean, standard deviation and minimum and maximum monthly mean were calculated.

  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al. 2012) for 32 fetch directions and coastline data at full resolution. Mean wave exposure (J m^-3^) at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

@@

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred. A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm (XXX).

  Beta diversity (sensu XXX) was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

@@

  Sequence data were successfully generated from 84 coral samples collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The dataset comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1. From 2003 to 2010, the north shore of St. John was characterized by lower wave exposure, slightly higher and more variable sea surface temperatures, and higher and more variable chlorophyll a concentrations, relative to the south shore. (Statistical tests?)

```{r Fig1, fig.height=3, fig.width=6, fig.cap="Figure 1. Sampling locations around St. John, U.S. Virgin Islands."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

pdf(file="figures/map.pdf", height=3, width=6)
layout(mat=matrix(c(1,1,6,2,1,1,5,3,1,1,5,4), nrow=3, byrow=T))

par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.75, -64.69), ylim=c(18.303, 18.370))
#box()
text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.77, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.77, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
})
#points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
#text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)
#dev.off()

# Plot environmental data
envdata <- read.csv("data/STJ_envdata.csv", check.names=F, row.names=1)
sites <- envdata[-1,4:5]
colnames(sites) <- c("North", "South")
sites$North <- as.numeric(as.character(sites$North))
sites$South <- as.numeric(as.character(sites$South))
sites <- data.frame(t(sites))

#SST
sst <- sites[, 2:6]
sst <- within(sst, {
  `Mean+S.D.` <- Mean.SST + S.D..SST
  `Mean-S.D.` <- Mean.SST - S.D..SST
})

par(cex=0.7, mar=c(2,0,1,2), mgp=c(1.5,0.5,0), xpd=NA)
plot(NA, xlim=c(26,29), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
axis(side=1, at=seq(26,29,1))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
bxp(list(stats=matrix(as.numeric(sst[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(sst[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

#chlA
chla <- sites[, 7:11]
chla <- within(chla, {
  `Mean+S.D.` <- Mean.Chla + S.D..Chla
  `Mean-S.D.` <- Mean.Chla - S.D..Chla
})

#par(new=T, mar=c(7,14,10,8), cex=0.7)
plot(NA, xlim=c(0.03,0.09), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
axis(side=1, at=seq(0.03,0.09,0.01))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
bxp(list(stats=matrix(as.numeric(chla[1, c(1,7,3,6,2)],5,1)), n=20), add=T, at=2, axes=F, horiz=T)
bxp(list(stats=matrix(as.numeric(chla[2, c(1,7,3,6,2)],5,1)), n=20), add=T, at=1, axes=F, horiz=T)

plot(NA, xlim=c(0,1), ylim=c(0.5,2.5), xlab="", ylab="", bty="n", yaxt="n")
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
#text(par("usr")[1], c(1,2), c("O","O"), cex=1.8, pos=2, offset=0.7, lwd=0.1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)

#### histogram of samples
nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
nsam <- dcast(nsam, Species~Shore)
rownames(nsam) <- nsam$Species

rownames(nsam) <- str_replace_all(rownames(nsam), 
                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
                  "siderea"="S.sid.", "strigosa"="P.str."))
nsam <- nsam[,2:3]
par(mar=c(4,0,6,4))
barplot(t(as.matrix(nsam)), las=2)
legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1.5, legend=c("N","S"), bty="n", inset=c(0.1,-0.2), xpd=T, y.intersp=0.8, x.intersp=0.5)
dev.off()
```

@@

   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n`.

```{r Table2}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

  A subset of samples was selected for comparative analysis of the three clustering approaches (Fig. 1). 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples. Because the outcomes of this approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach. Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.

@@

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore. Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
```

@@

  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any cnidarian species (Table 3). However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056). *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).

```{r Table3, cache=T } 
set.seed(43789) 
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 3. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

#### Network analyses of *Symbiodinium* meta-community
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

  `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs occurred as "dominant" symbionts in the corals sampled, defined by >50% relative abundance within a host (Fig. 5A). A single OTU identical to *Symbiodinium* B1 was the most prevalent dominant symbiont, dominating all *D. cylindrus* samples and many *D. strigosa* (4/9), *O. annularis* (2/4), *M. alcicornis* (6/10), and *F. fragum* (8/9) samples. Other closely-related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D. strigosa*, *M. alcicornis*, and *F. fragum*, while *O. annularis* was dominated just as often by *Symbiodinium* C7. After B1, the next most prevalent dominant symbionts were *Symbiodinium* C3, which dominated all *M. cavernosa* and most *S. siderea* samples (7/10), and *Symbiodinium* A4, which dominated all *P. astreoides* and 3/10 *P. furcata* samples. *P. furcata* and *S. siderea* were each occasionally dominated by other symbiont OTUs, including C45 or C45a (*P. furcata*), and C1 or D1 (*S. siderea*). All coral species were dominated by at least one *Symbiodinium* OTU that also dominated at least one other coral species, except for *S. radians*, which was exclusively dominated by C46.

  There were `r sum(V(bkgnet3)$type==2)` symbionts that occurred at background levels (<1% relative abundance) in at least three coral species (Fig. 5B). *Symbiodinium* D1a (which has been formally described as *S. trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1a"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa). C31, C46, and B1 were each detected at background levels in 4 host species.

@@

  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals (i.e., average distance to centroid) of each host species was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity. Significant differences in symbiotic flexibility were detected among host species (Fig. 6). The highest distance to centroid was found in *D. strigosa*, *P. furcata*, *M. alcicornis*, *S. siderea*, and *M. annularis*, and the lowest was found in *M. cavernosa*. Flexibility was also low in *D. cylindrus*, *S. radians*, and *P. astreoides*, and intermediate in *F. fragum*.

>>>>

  Symbiotic microalgae (*Symbiodinium* spp.) strongly influence the performance, fitness, and stress-susceptibility of their coral hosts. Therefore, describing the diversity and community structure of *Symbiodinium* assemblages in corals, as well as their metacommunities on the scale of whole reefs, has become a routine and essential goal in coral reef research. High-throughput sequencing of ITS2 nrDNA can describe these communities with unprecedented resolution, and the ease of such analyses facilitates studies across large spatial scales, yet the quantity of data generated and the high intragenomic variability at the ITS2 locus complicates parsing intragenomic versus interspecific variability, with this task requiring application of appropriate bioinformatic pipelines to achieve meaningful descriptions of diversity. Using samples of *Symbiodinium* collected from ten coral species on shallow reefs on the north and south shores of St. John, US Virgin islands, we demonstrate that clustering *Symbiodinium* ITS2 sequences at 97% similarity within, but not across, samples collapses sequence diversity that is more likely to be intragenomic, while maintaining diversity that is more likely interspecific. While this approach cannot resolve all *Symbiodinium* species (e.g., when closely related species co-occur in a sample, or when ITS2 fails to differentiate them), it enables broad-scale metabarcoding of the evolutionarily complex, yet commonly-used ITS2 locus to generate taxa-abundance data that is suitable for ecological analysis. We utilized this approach to create a comprehensive metacommunity network of host-symbiont associations in St. John, which revealed *Symbiodinium* lineages that occupy 'dominant' vs. 'background' niches, and coral hosts that are 'flexible' or 'specific' in their associations, though no differences were observed between north and south shore environments. These methods shed new light on important questions in coral symbiosis ecology, and demonstrate how application-specific bioinformatic pipelines can improve the analysis of metabarcoding data in microbial metacommunity studies.

@@

  A total of 84 tissue samples from scleractinian and milleporine corals were collected from seven sites around St. John, US Virgin Islands, USA (Fig. 1), between August 7th and 9th, 2012. Samples were collected on snorkel from 3-7m depth, and all colonies were selected haphazardly in upward facing locations exposed to ambient irradiance at the sampling depth. Tissue biopsies (~5 mm diameter and ~5 mm depth) were removed from coral colonies with sterile metal borers, and the small holes were plugged with non-toxic modeling clay.

  Broad-scale differences in the hydrographic regimes at points representative of the north and south shores of St. John (North: 18.368141°N, 64.711967°W; South: 18.315465°N, 64.726840°W) were evaluated from remote sensing tools to determine sea surface temperature (°C), surface chlorophyll concentrations (mg chl a m^-3^), and wave exposure (J m^-3^). These features were selected as they corresponded to anecdotal observations of differences between the water masses on both shores, and reflected conditions that strongly affect coral performance and *Symbiodinium* biology (i.e., SST [@Coles:1977p7350] and wave regime [@Atkinson:1994un]), and reflect nutrients and planktonic food resources that play important roles in coral nutrition [@Houlbreque:2009gh]. SST was evaluated from nighttime records, and chlorophyll from ocean colour data, both of which were obtained at 1 km spatial resolution from the MODIS sensor aboard the Aqua satellite for each month of the year, averaged over the period 2003-2010. We used as input the monthly climatologies available at the IMaRS website (http://imars.usf.edu/modis); if data were not available at the chosen coordinates, values from the nearest available pixel were used. Boxplots were generated from the 12 monthly climatological values obtained for each response (Fig. 1).

  Wind-driven wave exposure for a given site is dependent on the wind patterns and the configuration of the coastline, which defines the fetch, or the length of water over which a given wind has blown to generate waves. To calculate wave exposure, wind speed and direction data at each location were acquired from the QuickSCAT (NASA) satellite scatterometer from 1999 to 2008 at 25 km spatial resolution [@Hoffman:2005dz]. Coastline data were obtained from the Global Self-consistent, Hierarchical, High-resolution, Shoreline (GSHHS v 2.2) database, which provides global coastline at 1:250,000 scale [@Wessel:1996wk]. From these data, wave exposure was calculated using the methods based on wave theory (after Chollett et al. 2012) for 32 fetch directions and coastline data at full resolution. Mean wave exposure at each station was calculated in R using the packages maptools [@maptoolsToolsfor:_dYm01UK], raster [@rasterGeographicD:524kz2M0], rgeos [@rgeosInterfaceto:8WVz_nNk], and sp [@Bivand:2013ux].

@@

  Network analysis and visualization was performed in R using the igraph package [@Csardi:2006vh]. Networks were created featuring 'dominant' (>50% relative abundance) and 'abundant' (>1% relative abundance) OTUs, with weighted edges proportional to the number of individuals within a species in which a symbiont OTU occurred. A 'background' symbionts network was also created with unweighted edges defined based on the presence of a symbiont at <1% relative abundance in at least one individual within the species. To simplify visualization of the background symbiont network, clade D OTUs were merged into a single node under the assumption that clade D in the Caribbean comprises a single species (*Symbiodinium trenchii*; [@LaJeunesse:2014io]), and symbiont nodes connected to ≤ 2 coral species were removed from the network. All network layouts were constructed based on the Fruchterman-Reingold algorithm [@Fruchterman:1991ko].

  Beta diversity (sensu [@Anderson:2006fs]) was calculated as the multivariate dispersion of samples within a coral species. Principal coordinate analysis of Bray-Curtis dissimilarities of square-root transformed OTU counts was used to calculate average distance-to-centroid values for each species, which were then compared statistically by a permutation test. This analysis was implemented using betadisper() in the vegan package [@veganCommunityEco:uw], based on [@Anderson:2006fs].

@@

  A total of 84 coral samples representing ten different host species were collected across two sites on the north shore and five sites on the south shore of St. John (Fig. 1). The sequence data generated from these samples comprised `r system("grep -c '>' data/fasta/combined_seqs_trimmed.fasta")` sequences after merging paired reads, removing chimeric sequences, and trimming primers. After clustering, OTUs with <`r n` sequences and samples with <`r sn` sequences were excluded, leaving `r nsamples(phy97.f)` samples for downstream analysis.

  The environmental conditions broadly characterizing the north and south shores are presented in Figure 1. From 2003 to 2010, the north shore of St. John was characterized by slightly lower sea surface temperatures, higher chlorophyll a concentrations, and lower wave exposure, relative to the south shore.

```{r Fig1, fig.height=3, fig.width=6.85, fig.cap="Figure 1. Sampling of *Symbiodinium* communities from ten different host species (embedded histogram) on the north and south shores of St. John, US Virgin Islands, and the broad-scale physical conditions associated with these two environments. Mapped points along each shore indicate locations from which biological samples were collected, while the 'N' and 'S' points indicate locations from which environmental data representative of the north and south shores were obtained. Boxplots show the sea-surface temperatures (top), chlorophyll A concentrations (middle), and wave exposure (bottom) recorded at these two locations."}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

@@

#pdf(file="figures/map.pdf", height=3, width=6.85)
layout(mat=matrix(c(1,1,6,2,1,1,5,3,1,1,5,4), nrow=3, byrow=T))

# PLOT MAP
par(mar=c(0,0,0,0), oma=c(0,0,0,0), cex=1, xpd=NA)
plot(stjmap, lwd=1.5, col="black",
     xlim=c(-64.77, -64.70), ylim=c(18.303, 18.370))
#text(-64.735, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.78, -64.66, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.78, -64.66, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black", cex=0.8)
  #text(lon, lat, name, pos=c(4,2,4,4,4,1,3))
})
NS <- c(-64.711967, 18.368141)
SS <- c(-64.726840, 18.315465)
points(NS[1], NS[2], pch="N", cex=0.9, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch="S", cex=0.9, bg="gray40", col="black", lwd=1)
points(NS[1], NS[2], pch=1, cex=2.2, bg="gray80", col="black", lwd=1)
points(SS[1], SS[2], pch=1, cex=2.2, bg="gray40", col="black", lwd=1)

# Plot histogram of sample collections
par(new=T)
samdat <- data.frame(sample_data(phy97))
nsam <- with(samdat, aggregate(Species, by=list(Species=Species, Shore=Shore), FUN=length))
nsam <- dcast(nsam, Species~Shore)
rownames(nsam) <- nsam$Species
rownames(nsam) <- str_replace_all(rownames(nsam), 
                c("alcicornis"="M.alc.", "annularis"="O.ann.", "astreoides"="P.ast.","cavernosa"="M.cav.",
                  "cylindrus"="D.cyl.", "fragum"="F.fra.", "furcata"="P.fur.", "radians"="S.rad.",
                  "siderea"="S.sid.", "strigosa"="P.str."))
nsam <- nsam[,2:3]
par(mar=c(7,5,5,6), mgp=c(0.5,0.2,0))
barplot(t(as.matrix(nsam)), las=2, ylab="# samples", xpd=NA, cex.axis=0.5, cex.names=0.5, cex.lab=0.5, tcl=0.1)
legend("topleft", pch=22, pt.bg=c("gray90", "gray40"), pt.cex=1, cex=0.5, legend=c("N","S"), bty="n", inset=c(0.12,-0.1), xpd=T, y.intersp=0.8, x.intersp=0.75)

# Plot environmental data
envdata <- read.csv("data/STJ_envdata2.csv")
envdata$Site <- factor(envdata$Site, levels=c("S", "N"))

#SST
par(cex=0.7, mar=c(3,0,1,2), mgp=c(1,0.25,0), xpd=NA, cex.axis=0.7, cex.lab=0.7, tcl=-0.25)
plot(NA, xlim=c(25,30), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="SST (°C)", xaxt="n")
axis(side=1, at=seq(25,30,1))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
boxplot(envdata$SST ~ envdata$Site, range=0, add=T, axes=F, horizontal=T)

#chlA
plot(NA, xlim=c(0,0.1), ylim=c(0.5,2.5), bty="n", yaxt="n", ylab="", xlab="ChlA (mg m-3)", xaxt="n")
axis(side=1, at=seq(0,0.1,0.01))
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)
boxplot(envdata$chlA ~ envdata$Site, range=0, add=T, axes=F, horizontal=T)

#Wind exposure
we <- list(N=3.234787362,	S=4.025022708)
plot(NA, xlim=c(3,4.2), ylim=c(0.5,2.5), xlab="Wave exposure (J m-3)", ylab="", bty="n", yaxt="n")
points(c(we$N, we$S), c(2,1), pch=19)
text(par("usr")[1], c(1,2), c("S","N"), pos=2, offset=1)
points(rep(par("usr")[1]-strwidth("c.m"),2), c(1,2), pch=1, cex=2.5)

#dev.off()
```

@@

   The number of OTUs resolved, as well as the number of sequences per OTU and per sample, depended on the clustering approach used (Table 1 and Fig. S1). More OTUs were resolved as the clustering resolution increased (i.e., 97% OTUs across samples < 97% OTUs within samples < 100% OTUs), with fewer reads per OTU (Table 1). The number of sequences per sample was less for the 100% OTU approach because more OTUs were filtered out of the dataset by the minimum threshold count of `r n` reads per OTU.

```{r Table1}
# Compute summary statistics 
stats97.f <- data.frame(`97% OTUs<br>(across samples)`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% OTUs<br>(within samples)`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

@@

  A subset of samples was selected for comparative analysis of the community structure resolved by the three clustering approaches (Fig. 1). 97% across-sample clustering assigned the same dominant OTU to samples dominated by different sequence variants, while 97% within-sample clustering assigned distinct dominant OTUs to these samples. Because the outcomes of the latter approach are more consistent with the current understanding of ITS2 sequence diversity as it links to *Symbiodinium* biology and ecology (see Discussion), the remainder of the results is presented using the 97% within-sample clustering approach. Additional detailed comparisons of the outcomes of each clustering approach for each coral species are presented in the Supplementary Information.

@@

```{r Fig3, fig.height=6, fig.width=4, fig.align="center", fig.cap="Figure 3. *Symbiodinium* community composition of each sample. Samples are plotted as horizontal bars, sorted by species and shore of collection (north vs. south). Individual segments of each bar represent distinct 97% within-sample OTUs, colored by *Symbiodinium* clade identity."}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
```

@@

  No significant differences in *Symbiodinium* community composition between the north and south shores were detected for any host species (Table 2). However, some qualitative differences were apparent: *Porites furcata* colonies were dominated by either clade A or clade C *Symbiodinium* on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.056). *Siderastrea siderea* was dominated by either clade C or clade D on the south shore, but clade C only on the north shore (difference between shores PERMANOVA: p=0.071).

```{r Table2, cache=T } 
set.seed(43789)  
shorestats97bs <- perms(phy97bs.f.p, group="Species", trt="Shore")
knitr::kable(shorestats97bs, digits=3, row.names=F, 
             caption="Table 2. Bray-Curtis dissimilarities within and between shores for *Symbiodinium* communities in each coral species and PERMANOVA tests for a difference between shores.")

@@

#### Network analyses of *Symbiodinium* metacommunity
```{r networks,  cache=F, include=FALSE}
# Create dominant symbionts network
set.seed(12374)
domnet <- sppnet(phy=phy97bs.f.p, plot=F,
                 fun=function(x) length(x[x>0.5])/length(x), 
                 layout=layout.fruchterman.reingold)
# Calculate number of dominant symbionts connected to each species
dom.Nedges <- sapply(levels(data.frame(sample_data(phy97bs.f.p))$Species),
                     FUN=function(x) length(E(domnet)[from(x)]))
dom.Nedges <- sort(dom.Nedges, decreasing=T)

@@

  The dominant symbionts network contained `r sum(V(domnet)$type==2)` different *Symbiodinium* OTUs, defined as those occurring at >50% relative abundance within a host (Fig. 5A). A single OTU identical to *Symbiodinium* B1 was the most prevalent dominant symbiont, dominating all *D. cylindrus* samples and many *D. strigosa* (4/9), *O. annularis* (2/4), *M. alcicornis* (6/10), and *F. fragum* (8/9) samples. Other closely-related clade B OTUs (similar to B1, B1d, and B19) occasionally dominated *D. strigosa*, *M. alcicornis*, and *F. fragum*, while *O. annularis* was dominated just as often by *Symbiodinium* C7. After B1, the next most prevalent dominant symbionts were *Symbiodinium* C3, which dominated all *M. cavernosa* and most *S. siderea* samples (7/10), and *Symbiodinium* A4, which dominated all *P. astreoides* and 3/10 *P. furcata* samples. *P. furcata* and *S. siderea* were each occasionally dominated by other symbiont OTUs, including C45 or C45a (*P. furcata*), and C1 or D1 (*S. siderea*). All coral species were dominated by at least one *Symbiodinium* OTU that also dominated at least one other coral species, except for *S. radians*, which was exclusively dominated by C46.

  There were `r sum(V(bkgnet3)$type==2)` symbionts that occupied a 'background' niche, defined as <1% relative abundance in at least three coral species (Fig. 5B). *Symbiodinium* D1a (which has been formally described as *S. trenchii* [@LaJeunesse:2014io]) was found at background levels in the greatest number of host species examined (`r nspp["D1a"]`), followed by A4 and C3 (both were in `r nspp["C3"]` taxa). C31, C46, and B1 were each detected at background levels in 4 host species.

@@

  Symbiont communities from each individual host were ordinated in multivariate space, and the dispersion of individuals within each host species (i.e., average distance to centroid) was calculated as a metric of symbiosis 'flexibility'; more flexible host species will show greater variability in symbiont community composition among individuals, while less flexible host species will show greater uniformity. This metric revealed significant differences in symbiotic flexibility among host species (Fig. 6). The highest distance to centroid was found in *D. strigosa*, *P. furcata*, *M. alcicornis*, *S. siderea*, and *M. annularis*, and the lowest was found in *M. cavernosa*. Flexibility was also low in *D. cylindrus*, *S. radians*, and *P. astreoides*, and intermediate in *F. fragum*.
